---
title: "Effect-size visualizations for sequence analysis: design boundaries"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Effect-size visualization boundaries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

[DABEST estimation plots](https://acclab.github.io/DABEST-python/)
make sense for *scalar* outcomes extracted from ordered sequences,
such as participant-level transition probability, dwell fraction or
entropy under explicitly declared exposure denominators. They do not
justify treating transitions from the same participant as independent
observations.

Use the maintained R [dabestr package](https://acclab.github.io/dabestr/)
directly to show a Gardner–Altman or Cumming plot if a scalar
participant-level mean difference is the scientific target. For
a paired design use the same person in both groups and a paired
resampling contract.

For time-varying transitions, HMM state probabilities, scanpath
distances and repeated trials, retain the current
`plot_sequence_group_inference`, `plot_time_varying_sequence_model`,
and other estimator-specific uncertainty visualizations. Generic raw
mean differences cannot reproduce their temporal, compositional
or hierarchical estimands.

This article adds no speculative sequence estimator and does not modify
the published package public namespace.


## Rendered scalar sequence-effect illustration (synthetic)

For a prespecified **participant-level** transition-proportion estimand,
a paired effect plot can be useful. The individual transitions within
participants must **not** be treated as independent observations.

```{r gp3sequences-paired, fig.width=7.5, fig.height=4.5}
set.seed(20261009)
n <- 30L
latent <- stats::rbeta(n, 8, 12)
control <- stats::rbinom(n, size = 40, prob = latent) / 40
treatment <- stats::rbinom(
  n, size = 40, prob = pmin(latent + .10, .95)
) / 40
graphics::plot(
  NA, xlim = c(.8, 2.2), ylim = c(0, 1), xaxt = "n",
  xlab = "", ylab = "Synthetic AOI-to-AOI transition proportion",
  main = "Paired subject summaries; 40 transitions per condition"
)
graphics::axis(1, at = 1:2, labels = c("Control", "Treatment"))
graphics::segments(
  1, control, 2, treatment,
  col = grDevices::adjustcolor("#608B9A", alpha.f = .4)
)
graphics::points(rep(1, n), control, pch = 19, col = "#387C88")
graphics::points(rep(2, n), treatment, pch = 19, col = "#9265AA")
```

```{r gp3sequences-effect, fig.width=7.5, fig.height=4.5}
effect <- treatment - control
draws <- replicate(
  2000L, mean(sample(effect, size = length(effect), replace = TRUE))
)
interval <- stats::quantile(draws, c(.025, .975))
graphics::hist(
  draws, breaks = 25, col = "#8EB9C1", border = "white",
  main = "Bootstrap of participant-level transition effect",
  xlab = "Average difference in transition proportion"
)
graphics::abline(v = 0, lty = 3)
graphics::abline(v = mean(effect), col = "#245E69", lwd = 2)
graphics::abline(v = interval, col = "#B57D61", lty = 2, lwd = 2)
```

These are **illustrative synthetic outcomes**, not fitted sequence-model
results. For time-dependent transition probabilities, hidden states,
Markov dependence, exposure denominators or serially correlated repeated
trials, the scalar paired bootstrap may be the wrong estimand.
Use validated model-specific hierarchical and temporal uncertainty outputs.
