---
title: "gp3sequences: Transparent, Reproducible, and Auditable Analysis of Ordered Categorical Sequences in R"
abstract: >
  Ordered categorical sequences arise in longitudinal, behavioral, interaction, and process data, but reproducible analysis often requires researchers to coordinate incompatible representations, preprocessing choices, specialist methods, and reporting conventions. gp3sequences is an R package for transparent, reproducible, and auditable analysis of ordered categorical sequences stored in ordinary long-format data frames. It integrates explicit validation and preparation, descriptive summaries, contiguous and bounded noncontiguous pattern extraction, sequence dissimilarities, clustering and stability assessment, transition networks, higher-order and time-varying models, categorical hidden Markov models, declared-design inference, and guarded interoperability with specialist software. Common result structures retain settings, diagnostics, capabilities, provenance, and interpretation boundaries across analytical families. The package is positioned as an integration and audit layer rather than a replacement for specialist packages. A deterministic synthetic case study demonstrates the intended end-to-end workflow, while a primary-source-backed comparison clarifies overlap and delegation across seven established sequence-analysis packages. The software is designed to make analytical decisions inspectable and replayable without treating structural sequence outputs as evidence of psychological mechanisms or causal processes.
date: "2026-08-07"
author:
  - name: Stefanos Balaskas
    affiliation: "eGovernment & eCommerce Lab (Innovation & Entrepreneurship), Department of Business Administration, University of Patras"
    address: "26504 Patras, Greece"
    url: "https://sites.google.com/view/stefbalaskas/"
    orcid: "0000-0003-2444-9796"
    email: "s.balaskas@ac.upatras.gr"
output:
  rjtools::rjournal_article:
    toc: false
    self_contained: true
bibliography: gp3sequences.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.path = "figures/",
  fig.align = "center"
)

options(width = 80)

# Publication-only table formatting; analytical objects are unchanged.
journal_table <- function(x, caption = NULL, digits = 3L, ...) {
  x <- as.data.frame(x, check.names = FALSE)
  row.names(x) <- NULL

  # FINAL PUBLICATION DISPLAY TRANSFORMS
  # These alter printed presentation only; source evidence remains unchanged.

  normalise_display_name <- function(z) {
    z <- gsub("[^A-Za-z0-9]+", "_", z)
    z <- gsub("^_+|_+$", "", z)
    tolower(z)
  }

  # Table 2: concise three-column publication positioning.
  if (identical(
    caption,
    "Primary-source-backed positioning of gp3sequences and specialist comparators."
  )) {

    package_col <- which(
      normalise_display_name(names(x)) == "package"
    )

    stopifnot(length(package_col) == 1L)

    compact_comparison <- data.frame(
      Package = c(
        "gp3sequences",
        "TraMineR",
        "seqHMM",
        "arulesSequences",
        "WeightedCluster",
        "ClickClust",
        "PST",
        "march"
      ),
      Specialist.role = c(
        "Integrated auditable sequence workflows",
        "State sequences, visualization, and dissimilarities",
        "Categorical and mixture hidden Markov models",
        "Frequent sequential-pattern mining",
        "Clustering and typology validation",
        "Finite-mixture Markov clustering",
        "Probabilistic suffix trees",
        "Markov-family models"
      ),
      Positioning.relative.to.gp3sequences = c(
        "Focal integration and audit layer; structural outputs are not psychological or causal evidence",
        "Specialist state-sequence environment; no equivalent cross-family audit layer documented",
        "Specialist latent-state engine; not a general motif, distance, or network workflow",
        "Specialist pattern-mining engine; not a general sequence-analysis workflow",
        "Specialist clustering engine focused on clustering and typology validation",
        "Specialist model-based clustering engine focused on finite-mixture Markov clustering",
        "Specialist variable-memory engine focused on suffix-tree models",
        "Specialist Markov modelling engine focused on Markov-family models"
      ),
      stringsAsFactors = FALSE,
      check.names = FALSE
    )

    names(compact_comparison) <- c(
      "Package",
      "Specialist role",
      "Positioning"
    )

    package_values <- as.character(
      x[[package_col]]
    )

    order_idx <- match(
      package_values,
      compact_comparison$Package
    )

    stopifnot(
      !anyNA(order_idx),
      length(order_idx) == nrow(x)
    )

    x <- compact_comparison[
      order_idx,
      ,
      drop = FALSE
    ]
  }

  # TABLE 1 DISPLAY ROW EXPANSION
  # Display only: underlying workflow CSV remains unchanged.
  if (identical(
    caption,
    "Integrated evaluated case-study workflow."
  )) {

    nms <- normalise_display_name(names(x))

    stage_col <- which(nms == "stage")
    step_col <- which(nms == "manuscript_step")
    function_col <- which(nms == "gp3sequences_functions")

    stopifnot(
      length(stage_col) == 1L,
      length(step_col) == 1L,
      length(function_col) == 1L
    )

    function_parts <- strsplit(
      as.character(x[[function_col]]),
      ";[[:space:]]*",
      perl = TRUE
    )

    function_parts <- lapply(
      function_parts,
      trimws
    )

    stopifnot(
      all(lengths(function_parts) >= 1L),
      all(vapply(
        function_parts,
        function(z) all(nzchar(z)),
        logical(1L)
      ))
    )

    source_row <- rep(
      seq_len(nrow(x)),
      lengths(function_parts)
    )

    continuation <- unlist(
      lapply(
        lengths(function_parts),
        function(n) seq_len(n) > 1L
      ),
      use.names = FALSE
    )

    expanded <- x[
      source_row,
      ,
      drop = FALSE
    ]

    expanded[[function_col]] <- unlist(
      function_parts,
      use.names = FALSE
    )

    expanded[[stage_col]] <- as.character(
      expanded[[stage_col]]
    )

    expanded[[step_col]] <- as.character(
      expanded[[step_col]]
    )

    expanded[[stage_col]][continuation] <- ""
    expanded[[step_col]][continuation] <- ""

    row.names(expanded) <- NULL
    x <- expanded
  }

  # Table 5: turn internal policy tokens into reader-facing labels.
  if (identical(
    caption,
    "Explicit preparation decisions applied before analysis."
  )) {

    nms <- normalise_display_name(names(x))

    step_col <- which(nms == "step")
    policy_col <- which(nms == "policy")

    stopifnot(
      length(step_col) == 1L,
      length(policy_col) == 1L
    )

    x[[step_col]] <- tools::toTitleCase(
      gsub(
        "_",
        " ",
        as.character(x[[step_col]]),
        fixed = TRUE
      )
    )

    x[[policy_col]] <- tools::toTitleCase(
      gsub(
        "_",
        " ",
        as.character(x[[policy_col]]),
        fixed = TRUE
      )
    )
  }

  # Tables 8 and 9: shorten technical display headings.
  if (caption %in% c(
    "Most prevalent contiguous motifs within each synthetic group.",
    "Most prevalent bounded noncontiguous subsequences."
  )) {

    nms <- normalise_display_name(names(x))

    replacements <- c(
      motif_length = "Length",
      subsequence_length = "Length",
      n_occurrences = "Occurrences",
      occurrence_count = "Occurrences",
      sequence_count = "Sequences",
      n_sequences = "Sequences",
      sequence_prevalence = "Prevalence"
    )

    for (key in names(replacements)) {
      hit <- which(nms == key)

      if (length(hit) == 1L) {
        names(x)[[hit]] <- replacements[[key]]
      }
    }
  }

  # Table 14: compact inferential summary for the article.
  if (identical(
    caption,
    "Declared observational contrast in sequence-level prevalence of state A."
  )) {

    nms <- normalise_display_name(names(x))

    pick_one <- function(pattern) {
      hit <- grep(pattern, nms, perl = TRUE)

      if (length(hit) != 1L) {
        stop(
          paste0(
            "Could not uniquely identify inference column: ",
            pattern
          ),
          call. = FALSE
        )
      }

      hit
    }

    group1_col <- pick_one("^group_?1$")
    group2_col <- pick_one("^group_?2$")
    mean1_col <- pick_one("^mean(_group)?_?1$")
    mean2_col <- pick_one("^mean(_group)?_?2$")
    diff_col <- pick_one(
      "^difference(_group)?_?2_minus(_group)?_?1$|^difference$"
    )
    p_col <- pick_one("^p(_value)?$")
    permutation_col <- pick_one("^(n_?)?permutations$")
    lower_col <- pick_one("^(ci_)?lower(_bound)?$")
    upper_col <- pick_one("^(ci_)?upper(_bound)?$")

    lower_value <- as.numeric(x[[lower_col]])
    upper_value <- as.numeric(x[[upper_col]])

    x <- data.frame(
      "Group 1" = x[[group1_col]],
      "Group 2" = x[[group2_col]],
      "Mean 1" = x[[mean1_col]],
      "Mean 2" = x[[mean2_col]],
      "Difference (2-1)" = x[[diff_col]],
      "p" = x[[p_col]],
      "Permutations" = x[[permutation_col]],
      "95% CI" = sprintf(
        "[%.3f, %.3f]",
        lower_value,
        upper_value
      ),
      check.names = FALSE,
      stringsAsFactors = FALSE
    )
  }

  pretty <- c(
    sequence_id = "Sequence",
    position = "Position",
    state = "State",
    group = "Group",
    channel = "Channel",
    weight = "Weight",
    gp3sequences_functions = "Functions",
    from_state = "From",
    to_state = "To",
    n_transitions = "Transitions",
    origin_total = "Origin total",
    origin_transition_proportion = "Origin prop.",
    n_observations = "Observations",
    group_total = "Group total",
    proportion = "Proportion",
    prevalence = "Prevalence",
    n_sequences = "Sequences",
    n_pairs = "Pairs",
    avg_silhouette = "Avg. silhouette",
    average_silhouette = "Avg. silhouette",
    min_silhouette = "Min. silhouette",
    minimum_silhouette = "Min. silhouette",
    dunn_index = "Dunn index",
    within_between_ratio = "Within/between",
    singleton_count = "Singletons",
    out_degree = "Out deg.",
    in_degree = "In deg.",
    total_degree = "Total deg.",
    out_strength = "Out strength",
    in_strength = "In strength",
    total_strength = "Total strength",
    closeness = "Closeness",
    betweenness = "Between.",
    pagerank = "PageRank",
    n_permutations = "Permutations",
    n_boot = "Bootstraps",
    interpretation = "Interpretation",
    Primary.contribution = "Primary contribution",
    Relationship.to.gp3sequences = "Relationship to gp3sequences",
    Comparison.boundary = "Comparison boundary"
  )

  nm <- names(x)

  mapped <- ifelse(
    nm %in% names(pretty),
    unname(pretty[nm]),
    tools::toTitleCase(gsub('[._]+', ' ', nm))
  )

  names(x) <- mapped

  if (knitr::is_html_output()) {
    return(
      knitr::asis_output(
        as.character(
          knitr::kable(
            x,
            format = "html",
            caption = caption,
            digits = digits,
            row.names = FALSE,
            ...
          )
        )
      )
    )
  }

  tab <- knitr::kable(
    x,
    format = "latex",
    booktabs = TRUE,
    caption = caption,
    digits = digits,
    row.names = FALSE,
    escape = TRUE,
    ...
  )

  tab_lines <- strsplit(
    paste(as.character(tab), collapse = '\n'),
    '\n',
    fixed = TRUE
  )[[1L]]

  tab_start <- which(
    startsWith(tab_lines, "\\begin{tabular}")
  )

  tab_end <- which(
    tab_lines == "\\end{tabular}"
  )

  stopifnot(
    length(tab_start) == 1L,
    length(tab_end) == 1L
  )

  # FINAL TABLE FLOAT PLACEMENT
  table_env <- which(
    tab_lines == "\\begin{table}"
  )

  if (length(table_env) == 1L) {
    tab_lines[[table_env]] <- "\\begin{table}[!htbp]"
  }

  spec <- NULL

  if (identical(
    caption,
    "Integrated evaluated case-study workflow."
  )) {
    spec <- "p{0.05\\linewidth}p{0.22\\linewidth}p{0.53\\linewidth}"
  }

  if (identical(
    caption,
    "Primary-source-backed positioning of gp3sequences and specialist comparators."
  )) {
    spec <- "p{0.20\\linewidth}p{0.26\\linewidth}p{0.40\\linewidth}"
  }

  if (identical(
    caption,
    "Explicit preparation decisions applied before analysis."
  )) {
    spec <- "p{0.15\\linewidth}p{0.16\\linewidth}p{0.12\\linewidth}p{0.37\\linewidth}"
  }

  if (!is.null(spec)) {
    tab_lines[[tab_start]] <- paste0(
      "\\begin{tabular}{",
      spec,
      "}"
    )
  } else {
    needs_scale <-
      ncol(x) >= 6L ||
      sum(nchar(names(x))) > 70L

    if (isTRUE(needs_scale)) {
      tab_lines[[tab_start]] <- paste0(
        "\\resizebox{0.98\\linewidth}{!}{%\n",
        tab_lines[[tab_start]]
      )

      tab_lines[[tab_end]] <- paste0(
        tab_lines[[tab_end]],
        "\n}"
      )
    }
  }

  knitr::asis_output(
    paste(tab_lines, collapse = '\n')
  )
}
```

# Introduction

Ordered categorical sequences describe histories in which both the states
observed and their ordering are analytically relevant. Sequence-oriented
methods study whole trajectories rather than isolated observations and can
represent differences in ordering, timing, and duration
[@AbbottTsay2000; @StuderRitschard2016]. Although much of this methodological
literature developed around longitudinal social research, the representation
also applies to ordered behavioral paths, interaction traces, and event streams.

The R ecosystem already contains mature specialist software for major parts of
this problem. `TraMineR` provides state-sequence definition, description,
visualization, and dissimilarity analysis [@Gabadinho2011TraMineR], whereas
`seqHMM` provides hidden and mixture hidden Markov modelling for categorical
and multichannel sequences [@Helske2019seqHMM]. Other packages specialize in
frequent sequential patterns, clustering and typology validation, finite-mixture
Markov clustering, variable-memory models, or other Markov-family models.
This specialization is a strength of the R ecosystem rather than a deficiency.

A different problem arises when one applied analysis spans several of these
methodological families. Data may move between representations; preparation
choices may be encoded differently across methods; result objects expose
different diagnostics; and analytical settings can become difficult to compare
after multiple transformations. Sequence-analysis scholarship similarly
emphasizes complementarity between analytical approaches rather than a single
universally sufficient method [@RitschardStuder2018].

This article presents `gp3sequences` [@gp3sequences] as an integration and audit
layer for ordered categorical sequence workflows. Its contribution is not a new
universal sequence model and not a claim to replace specialist software.
Instead, the package provides a neutral long-format data contract, explicit
preparation policies, structured results, machine-readable diagnostics,
provenance records, interpretation boundaries, and guarded adapters that keep
heterogeneous analytical stages connected.

The contribution is evaluated in three ways. First, the article describes the
data and analysis contracts that organize the package. Second, it positions
`gp3sequences` against seven specialist packages using a primary-source-backed
comparison. Third, a deterministic synthetic case study exercises an end-to-end
workflow from validation and pattern extraction through dissimilarity,
clustering, transition networks, and predeclared observational inference.

# Design goals and scope

The package is organized around five design goals. First, input remains in
ordinary long-format data until a specialist representation is required.
Second, preparation decisions are explicit rather than embedded in downstream
model calls. Third, stochastic procedures expose seeds and relevant settings.
Fourth, analyses return structured diagnostics and provenance alongside their
principal numerical output. Fifth, interpretation is deliberately bounded:
structural patterns, clusters, communities, and latent states do not acquire
psychological, behavioral, or causal meaning from the software alone.

These goals separate workflow governance from methodological choice.
`gp3sequences` does not attempt to infer the correct distance, clustering
solution, transition order, or latent-state model automatically. Such choices
depend on the research question and on which characteristics of sequence
difference are substantively relevant. Comparative work on sequence
dissimilarities shows that measures vary in their sensitivity to sequencing,
timing, and duration, so method selection remains tied to the analytical
objective [@StuderRitschard2016]. The package therefore records analytical
choices rather than presenting them as universally appropriate defaults.

The public API spans validation and preparation, descriptive summaries,
contiguous motifs and bounded noncontiguous subsequences, dissimilarities,
clustering and stability assessment, transition networks, higher-order and
time-varying models, categorical hidden Markov workflows, declared-design
inference, and analysis auditing. Specialist engines remain appropriate when
their deeper model-specific functionality is the principal analytical need.

# Package architecture

## Data contract

The primary interface uses ordinary long-format data frames with explicit
sequence identifiers, within-sequence order, and categorical states. Optional
metadata remain attached at the observation or sequence level. This design
keeps analytical input inspectable and allows ordering, coding, missingness,
and preparation policies to be resolved before conversion to method-specific
objects.

The neutral representation is deliberately narrower than the object systems of
specialist packages. Its purpose is not to reproduce every specialist-object
attribute but to provide a stable boundary at which assumptions can be checked
and recorded. Conversion is postponed until a specialist representation is
actually required.

## Analysis contracts and provenance

Analysis results retain settings, capabilities, diagnostics, and interpretation
boundaries together with their principal numerical outputs. Provenance
utilities can inspect an individual result or compare related analyses without
assuming that outputs from distinct analytical families are interchangeable.
Choices such as distance normalization, clustering specification, transition
order, resampling counts, seeds, and declared comparison design therefore
remain available for later audit.

## Guarded delegation and interpretation

Interoperability is treated as explicit delegation. Adapters prepare inputs for
specialist packages where that delegation is useful, while dependency guards
keep optional software optional. This architecture is consistent with the
broader view that sequence analysis can profit from complementary longitudinal
methods rather than forcing every question through one model family
[@RitschardStuder2018].

The same boundary applies to interpretation. A cluster is reported as a
cluster, an HMM state as a model state, and a transition-network community as
a structural partition unless external theory and study design justify a
stronger interpretation.

# Integrated workflow

The intended workflow begins with validation and audit, proceeds through
explicit preparation and fit-for-purpose analysis, and ends with result
auditing and provenance comparison.

```{r workflow-plan}
workflow <- read.csv("data/case-study-workflow.csv", stringsAsFactors = FALSE)
journal_table(
  workflow[, c(
    "stage",
    "manuscript_step",
    "gp3sequences_functions"
  )],
  caption = "Integrated evaluated case-study workflow."
)
```

# Software comparison

The comparison is based on documented software purpose rather than raw
function counts. `TraMineR` provides the principal general state-sequence
environment in the comparison [@Gabadinho2011TraMineR], whereas `seqHMM`
specializes in hidden and mixture hidden Markov models
[@Helske2019seqHMM]. `arulesSequences` exposes frequent sequential-pattern
mining [@Buchta2026arulesSequences], including the cSPADE route whose
algorithmic foundation is SPADE [@Zaki2001SPADE]. `WeightedCluster` provides
dedicated clustering and cluster-quality tools
[@Studer2025WeightedCluster; @Studer2013WeightedCluster]. `ClickClust` focuses
on finite-mixture Markov clustering [@Melnykov2016ClickClust], `PST` on
probabilistic suffix trees and variable-length Markov models
[@Gabadinho2016PST], and `march` on Markov-family modelling
[@Maitre2020march; @BerchtoldMaitreEmery2020].

A capability classified as `not_documented` was not located in the reviewed
primary sources; this is not evidence that every historical or development
version lacks that capability. Likewise, `outside_scope` records that a
comparison dimension does not match the specialist purpose of a package.
The resulting matrix is therefore a positioning device rather than a league
table.

```{r software-comparison}
comparison <- read.csv("data/software-comparison.csv", check.names = TRUE)
journal_table(
  comparison[, c(
    "Package",
    "Primary.contribution",
    "Relationship.to.gp3sequences",
    "Comparison.boundary"
  )],
  caption = "Primary-source-backed positioning of gp3sequences and specialist comparators."
)
```

The comparison identifies a division of labour rather than a claim of
supremacy. When a study is centered on a specialist task, the corresponding
specialist package can provide a deeper and more direct interface. The
distinctive role of `gp3sequences` is instead the connection of preparation,
multiple analytical families, diagnostics, explicit design declarations, and
provenance within one inspectable workflow.

# Reproducible case study

## Executed workflow interface

The evaluated case-study script loads the repository source and invokes the
exported API directly. Validation preserves `analysis_unit` as metadata;
`compute_sequence_distance()` uses Levenshtein dissimilarity normalized by
maximum sequence length; `cluster_sequences()` applies the predeclared
two-cluster hierarchical solution; and
`declare_sequence_comparison_design()` explicitly identifies `analysis_unit`
as the independent-unit field. The subsequent
`test_sequence_group_difference()` call uses 999 permutations and fixed seed
`20260807`. In this synthetic design, `analysis_unit` is an explicit metadata
copy of `sequence_id`, making the independence declaration inspectable rather
than implicit.

## Data and declared design

The case study uses a deterministic synthetic data set with 72 sequences,
24 ordered observations per sequence, four categorical states, and two
declared groups. It provides a compact end-to-end example without
embedding private or proprietary data in the manuscript.

```{r case-data}
case_data <- read.csv("data/case-study-sequences.csv", stringsAsFactors = FALSE)
stopifnot(
  nrow(case_data) == 72L * 24L,
  length(unique(case_data$sequence_id)) == 72L,
  setequal(unique(case_data$state), c("A", "B", "C", "D"))
)

journal_table(
  utils::head(case_data, 12L),
  caption = "First observations in the deterministic synthetic case study."
)
```

## Validation and explicit preparation

The synthetic input was evaluated against the neutral long-format contract
before analysis. Preparation preserved all rows and consecutive repeats while
recording every resolved policy. Review-severity diagnostics are reported as
structural properties rather than silently removed.

```{r evaluated-validation}
validation_result <- read.csv("results/case-study/validation-summary.csv")
preparation_result <- read.csv("results/case-study/preparation-decisions.csv")
journal_table(
  validation_result,
  caption = "Validation status for the deterministic case-study data."
)
journal_table(
  preparation_result,
  caption = "Explicit preparation decisions applied before analysis."
)
```

## State and transition structure

The two declared groups were designed to differ structurally. Comparison
sequences alternate primarily between states A and C, whereas reference
sequences distribute observations across all four states. These differences
are properties of the deterministic generator and are not estimates of an
external population.

```{r evaluated-states, fig.align="center", fig.pos="!htbp", out.width="70%", fig.cap="Observation proportions by state and declared synthetic group.", fig.alt="Grouped bars compare observation proportions across states and the two declared synthetic groups."}
state_result <- read.csv("results/case-study/state-summary-by-group.csv")
journal_table(
  state_result,
  digits = 3,
  caption = "State frequencies and proportions by declared synthetic group."
)
knitr::include_graphics(
  "figures/case-study/state-distribution-by-group.png"
)
```

```{r evaluated-transitions}
transition_result <- read.csv("results/case-study/transition-summary-by-group.csv")
transition_top <- transition_result[
  order(
    transition_result$group,
    -transition_result$n_transitions,
    transition_result$from_state,
    transition_result$to_state
  ),
]
transition_top <- do.call(
  rbind,
  lapply(split(transition_top, transition_top$group), utils::head, 6L)
)
journal_table(
  transition_top,
  digits = 3,
  caption = "Most frequent adjacent transitions within each synthetic group."
)
```

## Recurring motifs and bounded subsequences

Contiguous motifs of lengths two and three and bounded noncontiguous
two-state subsequences were extracted using declared search limits. Reported
prevalence denotes structural recurrence across synthetic sequences; it is
not a significance test or a substantive mechanism.

```{r evaluated-patterns}
motif_result <- read.csv("results/case-study/motif-summary-by-group.csv")
subsequence_result <- read.csv("results/case-study/subsequence-summary-by-group.csv")
motif_top <- do.call(
  rbind,
  lapply(split(motif_result, motif_result$group), utils::head, 5L)
)
subsequence_top <- do.call(
  rbind,
  lapply(split(subsequence_result, subsequence_result$group), utils::head, 5L)
)
journal_table(
  motif_top,
  digits = 3,
  caption = "Most prevalent contiguous motifs within each synthetic group."
)
journal_table(
  subsequence_top,
  digits = 3,
  caption = "Most prevalent bounded noncontiguous subsequences."
)
```

## Dissimilarity and descriptive clustering

Normalized Levenshtein dissimilarities were clustered hierarchically with a
predeclared two-cluster solution. The average silhouette was 0.803361,
showing strong descriptive separation under the deliberately structured
synthetic generator. This is a partition diagnostic, not evidence that two
latent populations exist. Typologies derived from clustering remain
descriptive constructions whose interpretation depends on the analytical
objective [@Studer2013WeightedCluster].

```{r evaluated-clustering, fig.align="center", fig.pos="!htbp", out.width="70%", fig.cap="Silhouette values for the predeclared hierarchical two-cluster solution.", fig.alt="Silhouette values for all synthetic sequences under the predeclared two-cluster solution."}
distance_result <- read.csv("results/case-study/distance-summary.csv")
cluster_result <- read.csv("results/case-study/cluster-validation.csv")
cluster_composition <- read.csv("results/case-study/cluster-composition.csv")
journal_table(
  distance_result,
  digits = 3,
  caption = "Pairwise normalized Levenshtein dissimilarity summary."
)
journal_table(
  cluster_result,
  digits = 3,
  caption = "Descriptive validation of the two-cluster solution."
)
journal_table(
  cluster_composition,
  digits = 3,
  caption = "Declared-group composition of the resulting clusters."
)
knitr::include_graphics(
  "figures/case-study/cluster-silhouette.png"
)
```

## Transition network and declared group contrast

A first-order directed transition network summarizes structural connectivity.
The group contrast was declared as observational and uses `analysis_unit` as
the independent-unit field. For sequence-level prevalence of state A, the
predeclared permutation analysis returned `p = 0.001`. Because the data are
deterministic and the design is observational, this value characterizes an
exchangeability-based association under the synthetic construction; it does
not establish a causal effect or population-level evidence.

```{r evaluated-network-inference, fig.align="center", fig.pos="!htbp", out.width="68%", fig.cap="Sequence-level prevalence of state A by declared synthetic group.", fig.alt="Bars compare sequence-level state-A prevalence between the two declared synthetic groups."}
centrality_result <- read.csv("results/case-study/transition-centrality.csv")
inference_result <- read.csv("results/case-study/group-inference.csv")
journal_table(
  centrality_result,
  digits = 3,
  caption = "Structural centrality in the pooled first-order transition network."
)
journal_table(
  inference_result,
  digits = 3,
  caption = "Declared observational contrast in sequence-level prevalence of state A."
)
knitr::include_graphics(
  "figures/case-study/state-a-prevalence-by-group.png"
)
```

## Reproducibility and interpretation boundary

All tables and figures in this section are regenerated by
`scripts/06-run-evaluated-case-study.R`. A manifest records file hashes,
row counts, package version, canonical content hashes, repository base commit, analysis-script hash, and the fixed analysis seed.
The evaluated workflow demonstrates package behavior under known synthetic
structure; it does not supply empirical evidence about people, behavior, or
causal processes.

# Discussion

The contribution of `gp3sequences` is primarily architectural. Specialist
sequence methods already exist in mature R packages; the practical problem
addressed here is keeping heterogeneous analytical stages connected to their
input assumptions, resolved preparation policies, diagnostics, declared
design, and provenance. A common long-format boundary and structured result
contracts reduce the amount of implicit state that has to be reconstructed
when an analysis crosses methodological families.

The software comparison also identifies when this integration layer should not
be the endpoint. `TraMineR`, `seqHMM`, `arulesSequences`, `WeightedCluster`,
`ClickClust`, `PST`, and `march` each provide specialist depth that would be
counterproductive to duplicate incompletely. `gp3sequences` therefore treats
interoperability and delegation as part of the architecture rather than as
evidence that one interface should subsume the surrounding ecosystem.

The deterministic case study exercises this distinction. Validation, explicit
preparation, motifs, bounded subsequences, normalized Levenshtein
dissimilarities, clustering, transition-network summaries, and declared-design
inference are connected through one reproducible analysis. The strong
two-cluster separation and permutation result are expected consequences of the
known generator. Their role is to demonstrate coherent propagation of
settings, diagnostics, and interpretation boundaries, not superior statistical
power or scientific discovery.

A further implication concerns method choice. Sequence dissimilarity, cluster
specification, transition order, model complexity, and inferential design
remain scientific decisions. Different dissimilarity measures emphasize
different combinations of sequencing, timing, and duration
[@StuderRitschard2016], illustrating why an auditable workflow cannot
substitute for substantive justification. The package is designed to make
these choices visible and reproducible, not to remove them.

# Limitations

Several limitations follow directly from the package's scope. First, the
evaluated case study is deterministic and synthetic. It provides a controlled
integration test but no external-validity evidence, and its separation metrics
must not be generalized to empirical populations. Second, this article does
not present a computational benchmark against specialist packages;
consequently, no performance-superiority claim is made.

Third, an integration layer necessarily introduces conventions of its own.
Long-format normalization and structured result contracts can add conversion
work when an analysis already lives entirely within a specialist object
system. Optional adapters also depend on the availability and semantics of
external packages. For analyses requiring the deepest model-specific
functionality, a specialist interface may therefore be simpler and more
expressive.

Fourth, breadth does not imply automatic method selection. The package does not
decide which sequence distance is substantively appropriate, how many clusters
exist in an empirical population, what an HMM state means, or whether a group
contrast is causal. These conclusions require study design, theory,
sensitivity analysis, and domain evidence outside the software.

Finally, the software comparison is deliberately source-bounded. A
`not_documented` classification means that the capability was not established
in the reviewed source set; it is not proof of universal feature absence.
Software evolves, so this boundary should be retained when the comparison is
updated.

At manuscript preparation, the public CRAN release of `gp3sequences` is 0.1.0,
whereas the computational evidence reported here was generated from
development version 0.2.0.9000 at repository commit
`d863b95a9a05cc2ef8bad00c47a6d2d553dd0772`. These software states are
reported separately rather than presented as identical.

# Conclusion

`gp3sequences` provides an auditable integration layer for ordered categorical
sequence analysis in R. Its contribution is the connection of explicit data
preparation, heterogeneous analytical families, structured diagnostics,
guarded interoperability, declared-design inference, and provenance rather
than the replacement of specialist sequence-analysis software. The evaluated
synthetic workflow demonstrates that these contracts can be carried through
an end-to-end analysis while preserving explicit limits on interpretation.

# Reproducibility statement

The public package availability record is provided by the CRAN release
[@gp3sequences]. The numerical results in this article correspond to the
evaluated development snapshot 0.2.0.9000 at commit
`d863b95a9a05cc2ef8bad00c47a6d2d553dd0772`. Manuscript data, comparison
evidence, scripts, generated tables, figure sources, and case-study outputs are
stored under the repository's excluded `paper/` workspace.

The authoritative case-study script uses fixed seed `20260807` and records 20
frozen outputs with content hashes and provenance metadata. The
render-and-check workflow rebuilds the article from current manuscript source,
regenerates the submission products, and records journal-check evidence rather
than treating a previously rendered PDF as proof of reproducibility.

# Acknowledgements

The author thanks the developers and maintainers of R and of the specialist sequence-analysis packages considered in this article.
